The effects of context inference on motor learning: savings, de-adaptation and spontaneous recovery

### 2021-10-01 adaptation/article/article.tex
Third draft. Got rid of the section on volatility

\abstract{Abstract}
Humans have been shown to adapt their movements when a sudden change to the dynamics of the environment is introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Human are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. In this work, we use a computational mechanistic model to explain the behavioral phenomena effected by imperfect contextual information. Specifically, we present a hierarchical computational model for motor adaptation based on exact Bayesian inference. This model explicitly takes into account contextual information and how the dynamics of context inference affect learning and action selection. We show how the proposed model provides a unifying explanation for four different experimentally established phenomena: (i) effects of sensory cues and proprioceptive information on switching between tasks, (ii) the effects of environmental volatility and predictability on learning, (iii) the effects of previously-learned adaptations on switching between tasks, in addition to  (iv) the well studied savings, de-adaptation and spontaneous recovery.

@@

While these general models of adaptation explain the most common phenomena observed in experiments, other known phenomena remain outside of their scope. For example, it is known that adaptation rate is reduced in situations where the environment is unstable and unpredictable \citep{Herzfeld_memory_2014}, or situations in which errors are small \citep{Marko_Sensitivity_2012} and slowly introduced \citep{Huang_Persistence_2009}. Learning new adaptations has also been found to depend on the history of adaptations learned \citep{Vaswani_Decay_2013}.

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. Intuitively, in mechanical arm experiments, one can think of one context in which the arm exerts no force and one in which it does. By inferring the current context, a participant can know how to counteract the mechanical arm and succeed in a reaching movement. In some experiments, the context is clearly inferable via sensory cues \cite[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Adaptive_1994,Ethier_Spontaneous_2008}. In many other experiments, the context is not directly observable, and context inference is probabilistic and takes the form of an evidence-accumulating process. In this work, we propose that the behavioural consequences of this slow process of evidence accumulation can be observed in behavior and can be used to explain the phenomena outlined above.

@@

In the model, context is a categorical variable $\zeta$, where the identification of a context is done through Bayesian inference:
\begin{equation}
q(\zeta | \vartheta_t, \vartheta_{t-1}, c_{t-1}) \propto p(\vartheta_t | \vartheta_{t-1}, c_{t-1})p(\zeta)
\label{eqn:context-inference-mini}
\end{equation}
where the prior distribution over contexts $p(\zeta)$ includes information from two sources: (1) an expectation of continuity, encoding the expectation that the context does not change from trial to trial, and (2) an overall expectation of observing any one context, if the context did change. Below, we discuss how these two sources of information effect different experimentally observed phenomena.

@@

In some experiments, adaptation blocks are repeated throughout the experiments. For example, \cite{Oh_Minimizing_2019} used an O-A-O-A-O-A-O-A sequence. To differentiate the repeating blocks, the authors named them $O_i$ and $A_i$, where $i = 1, 2, ...$. Their experiment can be described as $O_1-A_1-O_2-...$. We adopt this naming convention.

@@

\cite{Oh_Minimizing_2019} performed partially-cued experiments similar to \cite{Kim_Neural_2015}. The authors performed two experiments that differed in the size of the adaptation: in the first experiment, the angle of rotation was 20 degrees, while in the second experiment it was only 10. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, respectively, alongside simulations with our model. Participants in the first experiment (\fref{fig:oh-2019}B) first learned the adaptation in $A_1$. From the $A_1 - O_2$ transition onward, participants showed immediate switching (with a one-trial lag) between A and O (both ways), which can be seen in terms of the blue line (adaptation) closely following the switches in the black line (true rotation). In the second experiment (\fref{fig:oh-2019}C), switching happens more slowly, with adaptation (blue line) lagging behind the switches in the real context (black line), and slowly catching up. As can be seen in the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large (B), and slow, noisy switches when it is low (C). This difference is explained by our model in terms of the size of the adaptation: as the size of the adaptation is lower (10 degrees), it becomes more difficult to distinguish errors made by incorrectly inferring the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

@@

\subsection{Slower adaptation in volatile and unpredictable environments}
One central element in the proposed model is that adaptation is modulated by the uncertainty on the current context: a prediction error observed at trial $t$ will effect motor adaptation with a magnitude proportional to the size of the error, where the proportionality constant is related to the uncertainty over the context in which the error was observed: the higher the uncertainty, the lower the size of the adaptation.

Direct experimental evidence for the effects of context inference on the rate of adaptation is provided by \cite{Herzfeld_memory_2014}. The authors showed that the volatility of the environment, expressed in terms of unpredictable, stochastic transitions between A and -A (and vice versa), change the speed at which adaptation occurs.

\cite{Herzfeld_memory_2014} used an experimental paradigm in which the force that a mechanical arm exerts on the participant's hand can change from trial to trial between O, A and -A. Participants were divided into three groups for which the context transition from one trial to the next had different probabilities, depending on the group: for the first group, transitions were very uncommon, for the second group they were somewhat common and for the third group they occurred at almost every trial. Importantly, these transitions were uncued, and the participant could only infer that a transition had occurred after a motor error was observed.

\cite{Herzfeld_memory_2014} found that the volatility of the environment affects learning. The more volatile the environment, the less participants adapt their motor responses after observing an error. In \fref{fig:herzfeld-2014}, we show results obtained with our model in a paradigm matching that of \cite{Herzfeld_memory_2014}. We simulated the conditions of low volatility and maximum uncertainty ($z = 0.9$ and $z = 0.5$ in \cite{Herzfeld_memory_2014}) for thirty trials, repeating the experiment 300 times with random sequences of A and -A each time, following the transition probabilities from the experiment. In \fref{fig:herzfeld-2014}A we show the adaptation as a function of trial number, averaged across all 300 runs. It can be seen that the agent matches the findings from \cite{Herzfeld_memory_2014} where adaptation is slower in the more volatile environment ($z = 0.5$) than in the less volatile one. Additionally, \fref{fig:herzfeld-2014}B shows the sensitivity to errors, here defined as $\text{adaptation} / \text{error}$; the more volatile environment creates a lower sensitivity to errors than the predictable environment, a difference that is consistent across all trials.

\begin{figure}
\centering
\includegraphics{./figures/figure_3.png}
\caption{Learning under uncertainty about contexts. 30 trials were simulated in the task defined by \cite{Herzfeld_memory_2014} and repeated 300 times to estimate the mean effects. Two conditions were used, following the naming convention from \cite{Herzfeld_memory_2014}: $z=0.5$ is an unpredictable condition, where the context has a 0.5 probability of switching from trial to trial. $z=0.9$ is a stable condition, where the context has a 0.9 probability of staying the same from trial to trial. (A) Adaptation as a function of trial number, averaged across all runs. (B) Error sensitivity, binned into bins of 5 trials, averaged across all runs. The error and consequently sensitivity to error (y axis) is in arbitrary units. Shaded areas are SEM.}
\label{fig:herzfeld-2014}
\end{figure}

To account for volatility, our model includes a term in context inference which assumes a transition can occur from the previous context to all other possible contexts (see Methods). Effectively, this transition term gives the model uncertainty on the current context, regardless of observations. The higher the volatility, the higher the uncertainty. This higher uncertainty is what lowers the observed learning rate across all trials.

@@

Experimental evidence supporting this view can be found in experiments with context switching. For example, \cite{Davidson_Scaling_2004} reported an experiment in which participants had to switch from 3A to A in one group, with a block sequence A-3A-A-3A, and from -A to A in another group, with a block sequence A-(-A)-A-(-A). After A and 3A (or -A in the other group) had been learned in the first two blocks, the authors found that the switch from 3A to A was faster than that from -A to A. The authors interpreted this as evidence that switching between adaptations happens more quickly if it is in the same direction as the current adaptation (e.g. both counter-clockwise), and more slowly if they are in the opposite direction (e.g. clockwise to counter-clockwise).

@@

We simulated data with our model fitting the experimental structure in \cite{Davidson_Scaling_2004}. We show the results in \fref{fig:davidson-2004}, alongside the experimental results from \cite{Davidson_Scaling_2004}. It can be seen that the agent exposed to the -8 context shifts back to 4 more slowly than the one exposed to +8, qualitatively reproducing the data from \cite{Davidson_Scaling_2004}, shown in \fref{fig:davidson-2004}B. Note that in our simulations, the difference between the groups, as well as the variability within each group, is smaller than in the experimental data. However, the effect can be reproduced with any number of simulated runs (each one representing one participant), pointing towards a reliable effect. While we focused here on a specific, well-known experiment, we discuss experiments similar to that by \cite{Davidson_Scaling_2004} in the Discussion.

\begin{figure}
\centering
\includegraphics{./figures/figure_4.png}
\caption{Motor error when switching back to a previously learned adaptation. Following the block structure from \cite{Davidson_Scaling_2004}, the first 40 trials of the switch from +8 (yellow) or -8 (blue) to 4 is shown. (A) Simulations from our model, with the task parameters from \cite{Davidson_Scaling_2004}, running eight simulations per group, each simulation representing one participant. On the y axis, the error is shown in arbitrary units, related to centimeters through a monotonically increasing transformation with the same origin. The shaded area is the SEM across simulated runs (participants). (B) Data adapted from figure 4 by \cite{Davidson_Scaling_2004}, showing the error in the same trials as (A).}
\label{fig:davidson-2004}
\end{figure}

Further evidence for this effect of uncertainty about contexts can be found during error-clamp blocks in several experiments. We discuss these experiments in the following section. Additionally, we discuss experiments similar to that by \cite{Davidson_Scaling_2004} in the Discussion section.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{./figures/figure_5.png}
\caption{Adaptation during error clamp trials. (A) Simulated adaptation during the error-clamp trials for the three groups of participants in \cite{Vaswani_Decay_2013}, using the same colors. The groups differ in the sequence of adaptations: 1.1 performed an A-E sequence; 1.2, O-A-E; 1.3, (-A/2)-A-E; and 1.4, (-A)-E. Following \cite{Vaswani_Decay_2013}, group 1.4 is not shown in A and B, as their behavior is identical to group 1.1. The solid line is the average across 10 runs (i.e. a group of 10 simulated participants) and the shaded area represents the standard deviation. The vertical dashed line is the start of the error-clamp trials. (B) Corresponding experimental data adapted from figure 2C by \cite{Vaswani_Decay_2013}. (C) Simulations: Inference over the current context, where contexts are color coded: black for baseline, orange for the counter-clockwise force, purple for the clockwise force and brown for counter-clockwise force with half strength. The lines represent the posterior probability of each context in every trial, while the background color represents the true context. An olive-colored background represents error-clamp trials. As in (A), solid lines represent the average across all runs and shaded areas represent the standard deviation. (D) Simulations: Visualization of the lag before a change in context is detected by the agent during the E trials. Each line represents one run (10 runs per group).}
\label{fig:vaswani-2013}
\end{figure}

@@

Our model also reproduces the variability in the lag before adaptation starts to drop to zero after the E block starts. In the model, this lag is due to a period in which context inference has not ``caught on'' to the change of context; during this period, behavior remains consistent with that of the previously-observed block, as can be seen in the experimental data as well. To show this, we show in \fref{fig:vaswani-2013}D the drop in the inferred probability of the previous context for each simulated run (one line per run, 10 runs per group, all groups). It can be seen that the moment in which the drop starts, as well as the speed of the drops, depends on the run, reproducing the different lags observed in experimental data. This can also be seen directly in context inference, displayed as the shaded areas in \fref{fig:vaswani-2013}C.

@@

\cite{Oh_Minimizing_2019} posited that the difference between their two groups (~\fref{fig:oh-2019}B and C) can be explained by considering that the smaller visual rotation of 10 degrees caused participants to update their baseline forward model instead of creating a new forward model for the context. If such an account were true, it would predict that when participants walk away from the experiment they would have trouble adapting to the no-rotation reality, having to re-learn it. Our model instead would suggest that participants would immediately switch back to their baseline movements the moment they walk away from the computer screen, as that is a clear indication that the context has switched back to baseline. More generally, our model predicts that the observed difference between the two groups would disappear if clear sensory contextual cues were added to the experiment. We believe our model provides a more general and principled explanation for the observations by \cite{Oh_Minimizing_2019} and, to our knowledge, such an effect has not been observed in humans in tasks like that by \cite{Oh_Minimizing_2019}.

>>>>

\abstract{Abstract}
Humans have been shown to adapt their movements when a sudden change to the dynamics of the environment is introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Human are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. In this work, we use a computational mechanistic model to explain the behavioral phenomena effected by imperfect contextual information. Specifically, we present a hierarchical computational model for motor adaptation based on exact Bayesian inference. This model explicitly takes into account contextual information and how the dynamics of context inference affect adaptation and action selection. We show how the proposed model provides a unifying explanation for four different experimentally established phenomena: (i) effects of sensory cues and proprioceptive information on switching between tasks, (ii) the effects of previously-learned adaptations on switching between tasks, in addition to  (iii) the well-studied savings, de-adaptation and spontaneous recovery.

@@

While these general models of adaptation explain the most common phenomena observed in experiments, other known phenomena remain outside of their scope. For example, it is known that adaptation rate is reduced in situations where the environment is unstable and unpredictable \citep{Herzfeld_memory_2014}, or situations in which errors are small \citep{Marko_Sensitivity_2012} and slowly introduced \citep{Huang_Persistence_2009}. Action selection has also been found to depend on the history of adaptations learned \citep{Vaswani_Decay_2013,Davidson_Scaling_2004}.

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. Intuitively, in mechanical arm experiments, one can think of one context in which the arm exerts no force and one in which it does. By inferring the current context, a participant can know how to counteract the mechanical arm and succeed in a reaching movement. In some experiments, the context is clearly inferable via sensory cues \cite[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Adaptive_1994,Ethier_Spontaneous_2008}. In many other experiments, the context is not directly observable, and context inference is probabilistic and takes the form of an evidence-accumulating process. In this work, we propose that the behavioral consequences of this slow process of evidence accumulation can be observed in behavior and can be used to explain the phenomena outlined above.

@@

In the model, context is a categorical variable $\zeta$ which is inferred using Bayes' theorem:
\begin{equation}
q(\zeta | \vartheta_t, \vartheta_{t-1}, c_{t-1}) \propto p(\vartheta_t | \vartheta_{t-1}, c_{t-1})p(\zeta)
\label{eqn:context-inference-mini}
\end{equation}
where the prior distribution over contexts $p(\zeta)$ includes information from two sources: (1) an expectation of continuity, encoding the expectation that the context does not change from trial to trial, and (2) an overall expectation of observing any one context, if the context did change. Below, we discuss how these two sources of information effect different experimentally observed phenomena.

@@

In some experiments, adaptation blocks are repeated throughout the experiments. For example, \cite{Oh_Minimizing_2019} used an O-A-O-A-O-A-O-A sequence. To differentiate the repeating blocks, the authors named them $O_i$ and $A_i$, where $i = 1, 2, ...$. Their experiment can be described as $O_1$-$A_1$-$O_2$-.... We adopt this naming convention.

@@

\cite{Oh_Minimizing_2019} performed partially-cued experiments similar to \cite{Kim_Neural_2015}. The authors performed two experiments that differed in the size of the adaptation: in the first experiment, the angle of rotation was 20 degrees, while in the second experiment it was only 10. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, respectively, alongside simulations with our model. Participants in the first experiment (\fref{fig:oh-2019}B) first learned the adaptation in $A_1$. From the $A_1 - O_2$ transition onward, participants showed immediate switching (with a one-trial lag) between A and O (both ways), which can be seen in terms of the blue line (adaptation) closely following the switches in the black line (true rotation). In the second experiment (\fref{fig:oh-2019}C), switching happens more slowly, with adaptation (blue line) lagging behind the switches in the real context (black line), and slowly catching up. As can be seen in the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large (B), and slow, noisy switches when it is low (C). This difference is explained by our model in terms of the size of the adaptation: as the adaptation is smaller (10 degrees), it becomes more difficult to distinguish errors made by incorrectly inferring the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

@@

@@

Experimental evidence supporting this view can be found in experiments with context switching. For example, \cite{Davidson_Scaling_2004} reported an curl-force experiment in which participants had to switch from 3A to A in one group, with a block sequence A-3A-A-3A, and from -A to A in another group, with a block sequence A-(-A)-A-(-A). After A and 3A (or -A in the other group) had been learned in the first two blocks, the authors found that the switch from 3A to A was faster than that from -A to A. The authors interpreted this as evidence that switching between adaptations happens more quickly if it is in the same direction as the current adaptation (e.g. both counter-clockwise), and more slowly if they are in the opposite direction (e.g. clockwise to counter-clockwise).

@@

We simulated data with our model fitting the experimental structure in \cite{Davidson_Scaling_2004}. We show the results in \fref{fig:davidson-2004}, alongside the experimental results from \cite{Davidson_Scaling_2004}. It can be seen that the agent exposed to the -A context shifts back to A more slowly than the one exposed to 3A, qualitatively reproducing the data from \cite{Davidson_Scaling_2004}, shown in \fref{fig:davidson-2004}B. Note that in our simulations, the difference between the groups, as well as the variability within each group, is smaller than in the experimental data. However, the effect can be reproduced with any number of simulated runs (each one representing one participant), pointing towards a reliable effect. While we focused here on a specific, well-known experiment, we discuss experiments similar to that by \cite{Davidson_Scaling_2004} in the Discussion.

\begin{figure}
\centering
\includegraphics{./figures/figure_3.png}
\caption{Motor error when switching back to a previously learned adaptation. Following the block structure from \cite{Davidson_Scaling_2004}, the first 40 trials of the switch from 3A (yellow) or -A (blue) to A is shown(A) Simulations from our model, with the task parameters from \cite{Davidson_Scaling_2004}, running eight simulations per group, each simulation representing one participant. On the y axis, the error is shown in arbitrary units, related to centimeters through a monotonically increasing transformation with the same origin. The shaded area is the SEM across simulated runs (participants). (B) Data adapted from figure 4 by \cite{Davidson_Scaling_2004}, showing the error in the same trials as (A).}
\label{fig:davidson-2004}
\end{figure}

Further evidence for this effect of uncertainty about contexts can be found during error-clamp blocks in several experiments. We discuss these experiments in the following section. Additionally, we discuss experiments similar to that by \cite{Davidson_Scaling_2004} in the Discussion.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{./figures/figure_4.png}
\caption{Adaptation during error clamp trials. (A) Simulated adaptation during the error-clamp trials for the three groups of participants in \cite{Vaswani_Decay_2013}, using the same colors. The groups differ in the sequence of adaptations: 1.1 performed an A-E sequence; 1.2, O-A-E; 1.3, (-A/2)-A-E; and 1.4, (-A)-E. Following \cite{Vaswani_Decay_2013}, group 1.4 is not shown in A and B, as their behavior is identical to group 1.1. The solid line is the average across 10 runs (i.e. a group of 10 simulated participants) and the shaded area represents the standard deviation. The vertical dashed line is the start of the error-clamp trials. (B) Corresponding experimental data adapted from figure 2C by \cite{Vaswani_Decay_2013}. (C) Simulations: Inference over the current context, where contexts are color coded: black for baseline, orange for the counter-clockwise force, purple for the clockwise force and brown for counter-clockwise force with half strength. The lines represent the posterior probability of each context in every trial, while the background color represents the true context. An olive-colored background represents error-clamp trials. As in (A), solid lines represent the average across all runs and shaded areas represent the standard deviation. (D) Simulations: Visualization of the lag before a change in context is detected by the agent during the E trials. Each line represents one run (10 runs per group).}
\label{fig:vaswani-2013}
\end{figure}

@@

Our model also reproduces the variability in the lag before adaptation starts to drop to zero after the E block starts. In the model, this lag is due to a period in which context inference has not ``caught on'' to the change of context; during this period, behavior remains consistent with that of the previously-observed block, as can be seen in the experimental data as well. To show this, we show in \fref{fig:vaswani-2013}D the drop in the inferred probability of the previous context for each simulated run (one line per run, 10 runs per group, all groups). The start and speed of the drop dependd on the run (i.e. the participant), reproducing the different lags observed in experimental data. This can also be seen directly in context inference, displayed as the shaded areas in \fref{fig:vaswani-2013}C.

@@

\cite{Oh_Minimizing_2019} posited that the difference between their two groups (~\fref{fig:oh-2019}B and C) can be explained by considering that the smaller visual rotation of 10 degrees caused participants to update their baseline forward model instead of creating a new forward model for the context. If such an account were true, it would predict that when participants walk away from the experiment they would have trouble adapting to the no-rotation reality, having to re-learn it. Our model instead would suggest that participants would immediately switch back to their baseline movements the moment they walk away from the computer screen, as that is a clear indication that the context has switched back to baseline. More generally, our model predicts that the observed difference between the two groups would disappear if clear sensory contextual cues were added to the experiment.
